Unifying intra- and inter-specific variation in tropical tree mortality

### 2018-07-21 ms/manuscript.Rnw
Added more text to the results and fixed some figures

In the remaining stages, we then take the most predictive hazard function and growth rate measure from stage 2 and examine how species traits, wood density, light demand and maximum dbh, affected model predictive performance. First, we individually fit an effect of each trait on all parameters (stage 3), we then examine 2-trait combinations on all parameters (stage 4) as well as the effect of three traits on all parameters (stage 5). Lastly, we fit a model that had included a species random effect on each parameter (stage 6). This allowed us to 1) examine the maximum predictive capacity when accounting for species differences; and 2) provides a relative indication on how much of the total species variation is captured by these three traits. Because these models were computationally expensive to run (some requiring over 400 computing hours), we were unable to examine interactions among traits, or determine the optimal combination of traits with parameters.

@@

\subsection*{Mortality over time}
Comparing the three 5-year intervals between censuses from 1995 to 2010, we found average mortality rates declining over time. The highest proportion of tree death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_training_full, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_training_full, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_training_full, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding the term $\delta$) we observed a small increase in predictive skill (Fig. \ref{fig2}A). Table S1 gives individual census effects.

@@

\subsection*{Hazard functions}
Comparing the three hazard functions summarised in Fig. \ref{fig1}B, we found that the third function --- the function containing both growth-dependent and growth-independent hazard terms, i.e. $(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the baseline hazard or growth-dependent only functions (Fig. \ref{fig2}). Moreover, we found that predictive skill was higher when using stem diameter growth over stem-area growth (Fig. \ref{fig2}A). Table S1 gives a summary of hyper parameter estimates.

@@

\subsection*{Species traits}
Using the hazard function with highest predictive skill (i.e. $(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$), we examined how both the individual and combined effects of species' maximum dbh, light demand and wood density affected model accuracy. We found that when we examined each trait individually on all three parameters ($\alpha$, $\beta$ and $\gamma$), both wood density and light demand significantly improved predictive performance relative to the model with no traits. By contrast, we found only marginal performance gains when a species' maximum dbh was included on all three parameters (Fig.~\ref{fig2} Stage 3).

Additional predictive gains were also obtained when multiple traits were considered. Specifically, we found that when the effects of light demand and maximum dbh where included on all parameters, predictive accuracy was significantly higher than any of the 1-trait models (Fig.~\ref{fig2} Stage 4). Additional, albeit marginal, gains in predictive performance were also obtained when the effects of all three traits where included on $\alpha$, $\beta$ and $\gamma$ (Fig.~\ref{fig2} Stage 5). However, relative to fitting species random effects on each parameter, the performance gain from these traits was small, signifying that substantial inter-specific variation in mortality rates remains unexplained (Fig. \ref{fig2} Stage 6). See Figs. S3-S5 for species-level parameter estimates.

Close examination of the how wood three traits on $\alpha$, $\beta$ and $\gamma$ revealed varied and contrasting effects.

Including wood density as an effect on either growth-independent or growth-dependent parameters significantly improved model skill relative to a model without such effects (Fig. \ref{fig2} A,B). The most parsimonious model, with highest predictive skill, was that attributing the wood density effect to the growth-independent hazard term ($\gamma$; Fig. \ref{fig2} B). Specifically, we found that wood density was negatively correlated with a species' baseline mortality rate, $\gamma$ (Fig. \ref{fig3}). This meant that fast growing individuals from a low wood density species had, on average, higher mortality rates (Fig. \ref{fig4} A,B), and thus higher probability of death across a 1-yr period (Fig. \ref{fig4} C,D), relative to fast growing individuals of high wood density species. For example, a species with a wood density of 0.3 g $\textrm{cm}^{-3}$ had an estimated mean probability of dying of \Sexpr{round(extract_baseline_estimate(final_model,FALSE,0.3),2)['mean']} yr$^{-1}$ compared to \Sexpr{round(extract_baseline_estimate(final_model,FALSE,0.8),2)['mean']} yr$^{-1}$ for a species with a wood density of 0.8 g $\textrm{cm}^{-3}$. Incorporating an additional species-level random effect to capture any additional inter specific differences substantially improved model skill relative to a model with only wood density (Fig. \ref{fig2}A). See Figs. S3-S5 for species-level parameter estimates.

To determine whether species-level parameters were associated with their maximum size or light requirement, we conducted a post-hoc analysis comparing the fitted parameters to these traits. We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the growth-independent hazard term, were weakly and positively correlated with gap index (a measure of species light requirement). By contrast, $\beta$ (the parameter that defines the exponential decay with increasing growth rate) was negatively correlated gap index. These correlations suggest that species which predominately recruit in gaps are more prone to death, due to both low growth and stochastic chance. They also required faster growth rates to achieve a given mortality rate relative to a species that recruit in shade. Correlations between estimated species parameters and their associated maximum DBH were weak).

\begin{figure}[h!]
\includegraphics{figures/fig4.pdf}
\caption{Predicted changes in annual mortality probability with increasing growth rate. A) Estimated average curves for all 203 Barro Colorado Island species used in this study. B,C,D) represent average ($\pm$ 95\% credible intervals) curve for a high (red) and low (blue) species trait values for wood density (B), maximum dbh (C), and light demand (D).}
\label{fig4}
\end{figure}

@@

<<results="asis", echo=FALSE>>=
Hmisc::latex(table_1(final_model, BCI_training_full),file = "", booktabs = TRUE, rowname = NULL,  
               colnamesTexCmd="bfseries",
               col.just = c(rep("c", 4)), label = "table:1",
               where = "!h",
               caption ="Relative effects of parameter removal on both r2. r2 estimates are derived 
               from modelled predictions vs the observed proportion of individuals that died between 1995 and 2010 for each species.")
@

>>>>

In the remaining stages, we then take the most predictive hazard function and growth rate measure from stage 2 and examine how species traits: wood density, light demand and maximum dbh, affected model predictive performance. Specifically, we first examined each trait individually on all three $\alpha$, $\beta$ and $\gamma$ (stage 3), we also examined 2-trait (stage 4) and 3-trait (stage 5) combinations on these three parameters. Lastly, we fit a model that included a species random effect on each parameter (stage 6). This allowed us to 1) examine the maximum predictive capacity when accounting for species differences; and 2) provides a relative indication on how much of the total species variation is captured by these three traits. Because these models were computationally expensive to run (some requiring over 400 computing hours), we were unable to examine interactions among traits, or determine the optimal combination of traits with parameters.

@@

\subsection*{Mortality over time}
Examining the three 5-year intervals between censuses from 1995 to 2010, we found average mortality rates declining over time. The highest proportion of tree death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_training_full, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_training_full, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_training_full, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding the term $\delta$) we observed a small increase in predictive skill (Fig. \ref{fig2}A). Table S1 gives individual census effects.

@@

\subsection*{Hazard functions}
Comparing the three hazard functions summarised in Fig. \ref{fig1}B, we found that the third function --- the function containing both growth-dependent and growth-independent hazard terms, i.e. $(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the baseline and growth-dependent only hazard functions (Fig. \ref{fig2}). Moreover, we found that predictive skill was higher when using stem diameter growth relative to stem-area growth (Fig. \ref{fig2}A). Table S1 gives a summary of hyper parameter estimates.

@@

\subsection*{Species traits}
Using the hazard function with highest predictive skill (i.e. $(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$), we examined how model predictive accuracy was influenced by adding the effects of species' maximum dbh, light demand and wood density to $\alpha$, $\beta$ and $\gamma$. In general we found that including traits effects on these parameters decreased logarithmic loss, and thus, increased model predictive accuracy (Fig.~\ref{fig2}). Moreover, we found that incorporating the effect of all three traits on these parameters resulted in better predictive model relative to models that only incorporated one or two of the traits. However, relative to a model including species random effects on each parameter, which account for unexplained species differences, the performance gain from these traits was small, signifying that substantial inter-specific variation in mortality rates remains unexplained (Fig. \ref{fig2} Stage 6). See Figs. S3-S5 for species-level parameter estimates.

Examining the effects of maximum dbh, light demand and wood density on $\alpha$, $\beta$ and $\gamma$ revealed that each trait influenced tree mortality in different ways (Fig. \ref{fig3}-\ref{fig4}). Specifically, we found that light demanding species were more susceptible to low growth rates (i.e. higher $\alpha$), had higher growth-independent hazards (i.e. higher $\gamma$) and exhibited shallower declines in growth-dependent hazards with increasing growth rates (i.e. lower $\beta$). By contrast, we found that wood density had significant impacts on $\alpha$ and $\gamma$, such that species with high wood density were less susceptible to low growth rates and had lower growth-independent hazards. Species' maximum dbh was found to only have a significant effect on $\gamma$, whereby species that can achieve larger dbh exhibited lower growth-independent hazards.

\begin{figure}[h!]
\includegraphics{figures/fig4.pdf}
\caption{Predicted changes in annual mortality probability with increasing growth rate. A) Estimated average curves for all 203 Barro Colorado Island species used in this study. B,C,D) represent average ($\pm$ 95\% credible intervals) curve for a high (red) and low (blue) trait values. B) wood density values 0.2 vs 0.8 (g cm$^{-3}$); C) maximum dbh values 3 vs 180 cm; D) light demand values 0.5 vs 0.8.}
\label{fig4}
\end{figure}

@@

<<results="asis", echo=FALSE>>=
Hmisc::latex(table_1(final_model, BCI_training_full),file = "", booktabs = TRUE, rowname = NULL,  
               colnamesTexCmd="bfseries",
               col.just = c(rep("c", 3)), label = "table:1",
               where = "!h",
               caption ="Relative effects of parameter removal on both r2. r2 estimates are derived 
               from modelled predictions vs the observed proportion of individuals that died between 1995 and 2010 for each species.")
@
